---
title: "Modelling growth"
author: 'eleanorjackson'
date: '`r format(Sys.time(), "%d %B, %Y")`' 
format: gfm+emoji
editor: source
---

```{r setup}
#| include: false

file_name <- knitr::current_input()

knitr::opts_chunk$set(
  fig.path =
    paste0("figures/", sub("\\.rmarkdown$", "", basename(file_name)), "/", sep = "")
)

ggplot2::theme_set(ggplot2::theme_classic(base_size = 10))
```

Whilst I'm waiting for the raw data, 
I thought I'd try fitting a few models to Ryan's cleaned data.

```{r packages}
#| output: false

library("tidyverse")
library("here")
library("patchwork")
library("janitor")
library("lubridate")
library("brms")

```

```{r}
read_csv(
  here::here("docs", 
             "r-veryard", 
             "data", 
             "combined_sbe_danum_data.csv")
  ) %>% 
  clean_names() -> ryan_data

glimpse(ryan_data)
```

Let's create a days-since-start variable. 
I'm not sure when the first measurements were taken, 
at the same time as planting out in the field,
or in the first survey weeks/months later?

Maybe it would be good to have the germination dates for each individual, 
then we could model by age - 
might allow us to tease out or account for the differences induced by different planting dates.

```{r}
min_date_primary <-
  ryan_data %>% 
  filter(type == "Primary") %>% 
  summarise(min_date = min(survey_date)) %>% 
  pluck(1,1)

min_date_secondary <-
  ryan_data %>% 
  filter(type == "Secondary") %>% 
  summarise(min_date = min(survey_date)) %>% 
  pluck(1,1)

ryan_data <-
  ryan_data %>% 
  mutate(days_since =
    case_when(type == "Primary" ~ difftime(survey_date, min_date_primary),
              type == "Secondary" ~ difftime(survey_date, min_date_secondary))
    ) %>% 
  mutate(days_since = time_length(days_since, unit = "days"))

```

```{r}
ryan_data %>% 
  ggplot(aes(x = days_since, y = log(dbh_mean))) +
  geom_point(alpha = 0.6) +
  geom_smooth(aes(colour = type),
              method = "glm", method.args = list(family = "gaussian")) +
  
  ryan_data %>% 
  ggplot(aes(x = days_since, y = dbh_mean)) +
  geom_point(alpha = 0.6) +
  geom_smooth(aes(colour = type),
              method = "glm", method.args = list(family = "gaussian")) +
  
  patchwork::plot_layout(guides = "collect")
```

I had a quick look at some other papers that model tree growth
and it seems that many of them just model rate of growth.

Let's try modelling tree growth rate (in mm per year)?

Growth rate = 

DBH at time 1 - DBH at time 2 / (difference in days / 365.25)

```{r}
first <- 
  ryan_data %>% 
  drop_na(dbh_mean) %>% 
  group_by(type, plot, tree_id, species) %>% 
  slice_min(days_since, n = 1) %>% 
  ungroup()

last <- 
  ryan_data %>% 
  drop_na(dbh_mean) %>% 
  group_by(type, plot, tree_id, species) %>% 
  slice_max(days_since, n = 1) %>% 
  ungroup()

growth_data <- 
  bind_rows(first, last) %>% 
  group_by(type, plot, tree_id, species) %>% 
  summarise(growth = max(dbh_mean) - min(dbh_mean),
            time = max(days_since) - min(days_since)) %>% 
  mutate(growth_rate = growth / (time / 365.25) ) %>% 
  drop_na(growth_rate)
```

```{r}
growth_data %>% 
  ggplot(aes(x = type, y = growth_rate)) +
  geom_boxplot()
```

```{r}
growth_data %>% 
  ggplot(aes(x = type, y = growth_rate)) +
  geom_boxplot() +
  facet_wrap(~ species)
```

Try a quick `lmer`

```{r}
m1 <- 
  lme4::lmer(
  formula = "growth_rate ~ type * species + (1 | plot)",
  data = growth_data
)

summary(m1)
```


```{r}
#| include: false

m2 <-
  brm(data = growth_data, 
      family = gaussian,
      scale(growth_rate) ~ type + species + (1 | plot),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 123)
```

Let's include an interaction between forest type and species.

```{r}
#| output: false
m3 <-
  brm(data = growth_data, 
      family = gaussian,
      scale(growth_rate) ~ type * species + (1 | plot),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 123)
```

```{r}
m3
```

Plotting the posterior estimates

```{r}
bayesplot::mcmc_areas(m3, pars = c("b_Intercept", "b_typeSecondary") )

```

```{r}
bayesplot::mcmc_areas(m3, regex_pars = "b_species") +
  bayesplot::mcmc_areas(m3, regex_pars = "b_typeSecondary:species") 

```

This seems promising. 

Seedlings grew slower in the primary forest compared to the secondary forest.
Species grow differently in the different forest types, see _S. macrophylla_.

## Conditional effects

```{r}
conditional_effects(m3) 
```

```{r}
conditional_preds <- marginaleffects::predictions(
  m3, 
  newdata = marginaleffects::datagrid(type = c("Primary", "Secondary")), 
  by = "type", 
  re_formula = NA
) %>% 
  marginaleffects::posteriordraws()

p_conditional_preds <- conditional_preds %>% 
  ggplot(aes(x = draw, fill = factor(type))) +
  ggdist::stat_halfeye() +
  labs(x = "Growth rate (scaled)", y = "Density", fill = "Forest type",
       title = "Conditional means") +
  theme(plot.subtitle = ggtext::element_markdown())

p_conditional_preds
```


If we do want to model growth specifically (rather than rate of growth),
I started reading 
[Applied Longitudinal Data Analysis: Modeling Change and Event Occurrence](https://solo.bodleian.ox.ac.uk/permalink/44OXF_INST/101jqdh/cdi_proquest_miscellaneous_37690487)
and Chapter 5 looked like it could be helpful.
They talk about how measurement collection is often not evenly spaced and 
how individuals can be different ages at the start of an experiment.
Thankfully, there is a {brms} and {tidyverse} translation of the book [avalible here.](https://bookdown.org/content/4253/)
