---
title: "A summary of the data"
author: 'eleanorjackson'
date: '`r format(Sys.time(), "%d %B, %Y")`' 
format: gfm+emoji
editor: source
knitr:
  opts_chunk: 
    code-fold: true
execute:
  warning: false
---

```{r setup}
#| include: false

file_name <- knitr::current_input()

knitr::opts_chunk$set(
  fig.path =
    paste0("figures/", sub("\\.rmarkdown$", "", basename(file_name)), "/", sep = "")
)

set.seed(123)
ggplot2::theme_set(ggplot2::theme_classic(base_size = 10))
```

I'm creating a document to summarise the clean dataset which we are working with
for the current manuscript.

Here we will look at:

- How many individual plants are there?
- How many species and how many individuals of each?
- How many plots are in each forest type?
- How frequently were the seedlings surveyed?

In future docs I'll do descriptive summaries of growth and survival.

```{r packages}
#| output: false

library("tidyverse")
library("here")
library("patchwork")
```

```{r}
# reading in data and adding a column distinguishing cohorts 1 & 2 from
# the primary forest seedlings
data <- 
  readRDS(here::here("data", "derived", "data_cleaned.rds")) %>% 
  mutate(census_no = as.ordered(census_no)) %>% 
  mutate(cohort = paste(forest_type, old_new, sep = "_")) %>% 
  filter(! cohort == "secondary_NA") %>% 
  filter(! str_detect(plant_id, "NA")) 
```

A glimpse of the data.

```{r}
glimpse(data)
```

## Individual plants & species

```{r}
data %>% 
  group_by(cohort, genus_species) %>% 
  summarise(n_individuals = n_distinct(plant_id)) %>% 
  pivot_wider(names_from = cohort, 
              values_from = n_individuals) %>% 
  rowwise() %>% 
  mutate(total = sum(primary_NA, 
                     secondary_N, 
                     secondary_O, 
                     na.rm = TRUE)) %>% 
  arrange(-total) %>% 
  janitor::adorn_totals() %>% 
  knitr::kable(col.names = c("Species", 
                             "Primary", 
                             "Secondary, 2nd cohort",
                             "Secondary, 1st cohort",
                             "Total # of individuals"))
```

There are `r length(unique(data$plant_id))` 
individual plants in the data from 
`r length(unique(data$genus_species))` species. 

`r length(unique(filter(data, forest_type == "primary")$plant_id))` 
in the primary forest, and 
`r length(unique(filter(data, forest_type == "secondary")$plant_id))` 
in the secondary forest.

38% of the 1^st^ cohort of secondary forest seedlings were replaced with a cohort 2 seedling.

```{r}
data %>% 
  group_by(forest_type, plot) %>% 
  summarise(n_species = n_distinct(genus_species)) %>% 
  pivot_wider(names_from = forest_type, 
              values_from = n_species) %>% 
  knitr::kable()
```

Plots contain either 14 or 15 different species.

Secondary forest plots were enrichment planted with 16 species each,
primary forest plots were planted with 23-25 species each.
Species were only included in this data if they occur in both the primary and secondary forest. 

```{r}
sp_lists <- 
  read_csv(here::here("data", "derived", "data_combined.csv")) %>% 
  select(forest_type, genus_species) %>% 
  distinct() %>% 
  drop_na(genus_species) %>% 
  group_by(forest_type) %>% 
  {setNames(group_split(., .keep = FALSE), group_keys(.)[[1]])}
```

Species in **both** the primary and secondary forest:

```{r}
intersect(sp_lists$secondary$genus_species, 
          sp_lists$primary$genus_species)
```

Species in the secondary forest which aren't in the primary forest:

```{r}
setdiff(sp_lists$secondary$genus_species, 
        sp_lists$primary$genus_species)
```

Species in the primary forest which aren't in the secondary forest:

```{r}
setdiff(sp_lists$primary$genus_species, 
        sp_lists$secondary$genus_species)
```

## Plots

```{r}
data %>% 
  group_by(forest_type) %>% 
  summarise(n_plots = n_distinct(plot)) %>% 
  knitr::kable()
```

There are 20 plots in the primary forest (plots `01:20`) and 
6 in the secondary forest (plots `03, 05, 08, 11, 14, 17`).

Only intensively sampled plots from the SBE (secondary forest) were included in this data,
and of the Danum plots (primary forest) only (sub-)plots in treefall gaps and those without drought treatments were included.

```{r}
data %>% 
  filter(forest_type=="secondary") %>% 
  group_by(plot) %>% 
  summarise(n_lines = n_distinct(line)) %>% 
  knitr::kable()
```

In the secondary forest there are 20 planting lines 
in each of the 6 intensively sampled plots.

```{r}
data %>% 
  filter(forest_type=="secondary") %>% 
  group_by(plot) %>% 
  summarise(n_positions = n_distinct(position)) %>% 
  knitr::kable()
```

The SBE was set up with the implicit assumption that each of the planting lines 
would have 66 positions where a seedling was planted.
However, due to some positions being un-plantable (i.e. obstructed by rock or stream),
and due to undulating topography (seedling was planted every 3m) there are often more or less than 66 planting positions.

There are more individual plants than positions as those that died
very early in the experiment were replaced by a new cohort of seedlings.

## Censuses

```{r}
data %>% 
  select(census_no, plot, forest_type) %>%
  distinct() %>% 
  ggplot(aes(x = census_no, y = plot)) +
  geom_point() +
  facet_wrap(~forest_type, scales = "free_x") +
  guides(x =  guide_axis(angle = 90)) 
```

There have been 22 censuses of the primary forest seedlings,
and 26 censuses of the secondary forest seedlings.

Plot 17 in the secondary forest was not sampled as frequently as the other
secondary forest plots as it wasn't included in surveys for the climber cutting study
(explained further below).

Not all the censuses were complete.

```{r}
data %>% 
  group_by(forest_type, census_no) %>% 
  summarise(n_distinct(plot)) %>% 
  pivot_wider(names_from = forest_type, 
              values_from = `n_distinct(plot)`) %>% 
  mutate(primary = (primary/20)*100,
         secondary = (secondary/6)*100) %>% 
  rename(`% of primary plots censused` = primary,
         `% of secondary plots censused` = secondary) %>% 
  knitr::kable(digits = 1)

```

Within the secondary forest, 
not every planting line was surveyed when the rest of the plot was censused.

Three of the secondary forest plots received a climber cutting treatment
(plots `05`, `11` and `14`) 
(see [O'Brien _et al_ 2019](https://doi.org/10.1111/1365-2664.13335)). 
When data was collected for this study only two planting lines from each of the
climber cut (plots `05`, `11` and `14`) and 
control plots (plots `03` and `08`) were surveyed 
since a full census would take too long.

```{r}
#| fig-width: 7
#| fig-height: 3

data %>% 
  mutate(month = lubridate::month(survey_date)) %>% 
  group_by(survey_date) %>% 
  slice_head() %>% 
  ggplot(aes(y = forest_type, x = survey_date, 
             colour = census_no)) +
  geom_point(alpha = 0.6, shape = "|", size = 10) +
  theme_bw() +
  scale_x_date(minor_breaks = "year") +
  scale_colour_viridis_d() +
  theme(legend.position = "none")
```

The censuses were not evenly spaced through time.

```{r}
#| fig-width: 7
#| fig-height: 2

data %>% 
  filter(forest_type == "secondary") %>% 
  mutate(census_type = str_extract(census_id, "[a-z]+" )) %>% 
  mutate(month = lubridate::month(survey_date)) %>% 
  group_by(survey_date) %>% 
  slice_head() %>% 
  ggplot(aes(y = forest_type, x = survey_date, 
             colour = census_type)) +
  geom_point(alpha = 0.6, shape = "|", size = 10) +
  theme_bw() +
  scale_x_date(minor_breaks = "year") 
```

Secondary forest plots which were included in each census type:

```{r}
data %>% 
  filter(forest_type == "secondary") %>% 
  mutate(census_type = str_extract(census_id, "[a-z]+" )) %>% 
  select(census_type, plot) %>% 
  distinct() %>% 
  pivot_wider(names_from = census_type, values_from = plot) %>% 
  knitr::kable()
```

Plots `05`, `11` and `14` received a climber cutting treatment.

- In the **full** surveys every individual in every plot was surveyed
- In the **intensive** surveys every individual in 6 of the plots were surveyed (all plots included in this data)
- In the **climber** surveys 2 planting lines of 5 plots were surveyed
