---
title: "Data, first look"
author: 'eleanorjackson'
date: '`r format(Sys.time(), "%d %B, %Y")`' 
format: gfm+emoji
editor: source
---

```{r setup}
#| include: false

file_name <- knitr::current_input()

knitr::opts_chunk$set(
  fig.path =
    paste0("figures/", sub("\\.rmarkdown$", "", basename(file_name)), "/", sep = "")
)

ggplot2::theme_set(ggplot2::theme_classic(base_size = 10))
```

In this notebook I'm going to take a first look at the Danum Gaps data.
Currently we've got 3 versions of the data: 

- some census-level excel files from Andy (which I think are raw data)
- the data published by Mikey [on Zenodo](https://zenodo.org/records/10701333) to accompany [his publication](https://doi.org/10.1038/s43247-024-01335-5)
- the cleaned data that Ryan used for his thesis [published on OSF](https://osf.io/cfb69/?view_only=812168fc288d4ffb89e7eec3dada01c1)

Do these data match up? Can we add newly collected data to Ryan's dataset?

```{r packages}
#| output: false

library("tidyverse")
library("here")
library("patchwork")
library("janitor")
```

Reading in the `DanumGaps_Data_*` files from Andy. 
There are 8 files: 2015, 2015_plus, 2016, 2017, 2018, 2019, 2023 and 2024.
Are these just the "intensive" plots? 
We should be able to add to the 2024 data this August.

```{r}
file_names <-
  as.list(dir(path = here::here("data", "raw"),
              pattern = "DanumGaps_Data_*", 
              full.names = TRUE))

data_list <-
  lapply(X = file_names, 
         FUN = readxl::read_excel,
         range = readxl::cell_cols("A:K"),
         na = c("", "NA"),
         col_types = c("date", 
                       "text",
                       "text",
                       "text",
                       "text",
                       "numeric",
                       "numeric",
                       "numeric",
                       "numeric",
                       "numeric",
                       "text") )

names(data_list) <-
  lapply(file_names, basename)

andy_data <-
  bind_rows(data_list, .id = 'df') %>%
  clean_names() %>% 
  mutate(across(canopy:plant_no, as.factor))

rm(data_list)
```

```{r}
glimpse(andy_data)

summary(andy_data)
```


```{r}
andy_data %>% 
  mutate(survey_yr = as.ordered(readr::parse_number(df))) %>%
  filter(canopy == "G") %>% 
  mutate(survival = as.factor(survival)) %>% 
  group_by(survey_yr, survival) %>% 
  summarise(n_obs = n()) %>% 
  ggplot(aes(x = survey_yr, y = n_obs, fill = survival)) +
  geom_col() 
```

I think there were two rounds of censusing in 2015, 
which is why the number of observations doubles.

Taking a look at the data from [Mikey on Zenodo](https://doi.org/10.5281/zenodo.10701333).
I don't think this is all the data that's been collected 
but may allow us to link plant ID's to species.

```{r}
read_tsv(
  here::here("data", "raw", "download_zenodo", "DataGrowthSurvivalClean.txt")
  ) %>% 
  clean_names() -> mikey_data

glimpse(mikey_data)
```

So, this data spans 2004 - 2017, and perhaps it includes more of the plots?

```{r}
n_distinct(mikey_data$plant_id)

n_distinct(andy_data$plant_no)
  
```

Yep ok, Mikey's data includes observations from 2,000 seedlings 
and Andy's data covers 36 seedlings. Is that right? Doesn't seem like many.

Let's check out the `plant_id` and `plant_no` variables 
to see if they are actually unique plant IDs or not.

```{r}
mikey_data %>% 
  group_by(block, canopy, plot, subplot,
           genus, species) %>% 
  summarise(n()) %>% 
  glimpse()
```

There are 1,932 unique seedlings when using block-canopy-plot-subplot 
as a unique identifier.

```{r}
mikey_data %>% 
  group_by(block, canopy, plot, subplot,
           genus, species, plant_id) %>% 
  summarise(n(), .groups = "drop") %>% 
  count()

n_distinct(mikey_data$plant_id)
```

`plant_id` is supposed to be a unique ID for individual seedlings.
When we include `plant_id` in our block-canopy-plot-subplot ID we get 2,017 counts,
although there are 2,000 distinct `plant_id` values.. :thinking:

I think the subplots might have different species abundances. 
They should have 25 seedlings each, but could be one of 26 different species.

```{r}
mikey_data %>% 
  group_by(block, canopy, plot, subplot) %>% 
  summarise(n_sp = n_distinct(species), 
            .groups = "drop") %>% 
  count(n_sp, name = "n_subplots")
```

Yep, this lines up!

Now taking a look at Andy's data.
We are missing the `plot` variable that's in Mikey's data.

```{r}
andy_data %>% 
  group_by(block, canopy, subplot,
           plant_no) %>% 
  summarise(n()) %>% 
  glimpse()
```

I think that in this data `plant_no` is only unique at the subplot level.
There is a max of 8 observations per row, 
which makes sense - one for each round of censusing. 
There are 284 unique rows which sounds like a more sensible number of seedlings.
But what species are they?

There is also the cleaned data that Ryan has made [available on OSF](http://bit.ly/3TKgKOo).

```{r}
read_csv(
  here::here("docs", 
             "r-veryard", 
             "data", 
             "combined_sbe_danum_data.csv")
  ) %>% 
  clean_names() -> ryan_data

glimpse(ryan_data)
```

Here it looks like `tree_id` is a combination of `plot`, `cohort` and 
perhaps a subplot level plant ID?

I'm just interested in the Danum plots at the moment.

```{r}
ryan_data %>% 
  filter(type == "Primary") -> ryan_data_danum

n_distinct(ryan_data_danum$tree_id) 
  
```

```{r}
ryan_data_danum %>% 
  mutate(year = year(survey_date)) %>% 
  group_by(year) %>% 
  summarise(n())
```

Ryan's data covers 2004 to 2017 - that's the same time span as Mikey's data.
2005, 2010, 2013, 2014, 2015 and 2017 have approx double the number of observations,
were two censuses done in those years?

I think we are going to need the raw data and scripts that Ryan used to 
create his `combined_sbe_danum_data.csv` file.
