---
title: "A summary of growth"
author: 'eleanorjackson'
date: '`r format(Sys.time(), "%d %B, %Y")`' 
format: gfm+emoji
editor: source
knitr:
  opts_chunk: 
    code-fold: true
execute:
  warning: false
---

```{r setup}
#| include: false

file_name <- knitr::current_input()

knitr::opts_chunk$set(
  fig.path =
    paste0("figures/", sub("\\.rmarkdown$", "", basename(file_name)), "/", sep = "")
)

set.seed(123)
ggplot2::theme_set(ggplot2::theme_classic(base_size = 10))
```

Here we will look at:

- Size of seedlings at their first survey
- Diameter measurements at BH vs basal 
- Growth between forest types
- Growth between species
- Interactions of species & forest types
- Effect of climber cutting on growth

```{r packages}
#| output: false

library("tidyverse")
library("here")
library("patchwork")
library("formattable")
```

```{r}
# reading in data and adding a column distinguishing cohorts 1 & 2 from
# the primary forest seedlings
data <- 
  readRDS(here::here("data", "derived", "data_cleaned.rds")) %>% 
  mutate(census_no = as.ordered(census_no)) %>% 
  mutate(cohort = paste(forest_type, cohort, sep = "_")) %>% 
  filter(! cohort == "secondary_NA") 
```

## Size of seedlings at their first survey

```{r}
data %>% 
  filter(survival == 1) %>% 
  group_by(plant_id) %>% 
  slice_min(survey_date, with_ties = FALSE) %>% 
  ggplot(aes(x = dbase_mean, 
             fill = cohort,
             colour = cohort)) +
  geom_density(alpha = 0.3) +
  facet_wrap(~genus_species, ncol = 5) +
  stat_summary(aes(xintercept = after_stat(x), y = 0), 
               fun = median, 
               geom = "vline", 
               orientation = "y",
               linetype = 2) +
  xlim(0, 20) +
  theme(legend.position = "top") +
  ylab("density") +
  xlab("basal diameter at first survey") 
```

```{r}
data %>% 
  filter(survival == 1 & 
           cohort != "secondary_2") %>% 
  group_by(plant_id) %>% 
  slice_min(survey_date, with_ties = FALSE) %>% 
  ggplot(aes(x = dbase_mean, 
             fill = cohort,
             colour = cohort)) +
  geom_density(alpha = 0.3) +
  facet_wrap(~genus_species, ncol = 5) +
  stat_summary(aes(xintercept = after_stat(x), y = 0), 
               fun = median, 
               geom = "vline", 
               orientation = "y",
               linetype = 2) +
  xlim(0, 20) +
  theme(legend.position = "top") +
  ylab("density") +
  xlab("basal diameter at first survey") 
```

Dashed line shows the median value.

Seedlings in the primary forest are usually smaller at the time of their first measurement.

The secondary forest of SBE was planted first then 
the left over seedlings in the nursery were planted in the primary old growth forest of Danum Valley.

Data suggests that the seedlings grew slower in the nursery 
than when planted out in the field and 
some species were more affected than others. 
The table below quantifies the difference in medians between forest types for each species.

```{r}
data %>% 
  filter(survival == 1) %>% 
  group_by(plant_id) %>% 
  slice_min(survey_date, with_ties = FALSE) %>% 
  group_by(genus_species, cohort) %>% 
  summarise(med = median(dbase_mean, na.rm = TRUE)) %>% 
  pivot_wider(names_from = cohort, values_from = med) %>% 
  select(-secondary_2) %>% 
  rowwise() %>% 
  mutate(difference_in_median_basal_diameter  = secondary_1 - primary_1) %>% 
  arrange(desc(difference_in_median_basal_diameter)) %>% 
  knitr::kable(digits = 2)
```

## Diameter measurements at BH vs basal

```{r}
data %>% 
  filter(survival == 1) %>% 
  group_by(cohort, census_no) %>% 
  summarise(n_dbh_only = sum(!is.na(dbh_mean) & is.na(dbase_mean)),
            n_basal_only = sum(!is.na(dbase_mean) & is.na(dbh_mean)),
            n_dbh_and_basal = sum(!is.na(dbh_mean) & !is.na(dbase_mean)) ) %>% 
  pivot_longer(cols = c(n_dbh_only, n_basal_only, n_dbh_and_basal), 
               names_to = "measurement",
               values_to = "n_plants") %>% 
  ggplot(aes(x = census_no, y = n_plants, 
             fill = measurement)) +
  geom_col(position = "fill") +
  facet_wrap(~cohort, ncol = 1,
             axis.labels = "all_x", axes = "all_x") +
  guides(x =  guide_axis(angle = 90)) +
  theme(legend.position = "top", legend.justification = "left") +
  geom_hline(yintercept = 0.5, colour = "white", linetype = 2) +
  geom_hline(yintercept = 0.75, colour = "white", linetype = 3)

```

At least 75% of seedlings have both DBH and basal diameter measurements by:

- Census 6 in the primary forest (2009) 
- Census 8 in the secondary forest (1^st^ cohort) (2012)

It takes longer before both diameter measurements are recorded for 
75% of the secondary forest's *2^nd^ cohort* of seedlings.

```{r}
data <-
  data %>%
  group_by(plant_id) %>%
  slice_min(survey_date, with_ties = FALSE) %>%
  select(plant_id, survey_date) %>%
  rename(first_survey = survey_date) %>%
  ungroup() %>% 
  right_join(data)


data <-
  data %>%
  rowwise() %>% 
  mutate(
    days =
      survey_date - first_survey) %>% 
  ungroup() %>% 
  mutate(years = as.numeric(days, units = "weeks")/52.25,
         days_num = as.numeric(days))
```

## Growth by forest type / cohort

```{r}
all_gr <- 
  data %>% 
  group_by(plant_id, genus_species, forest_type) %>%
  summarise(dbh_diff = max(dbh_mean, na.rm = TRUE) - 
              min(dbh_mean, na.rm = TRUE),
            basal_diff = max(dbase_mean, na.rm = TRUE) - 
              min(dbase_mean, na.rm = TRUE),
            max_years = max(years, na.rm = TRUE), 
            .groups = "keep") %>% 
  filter_if(~is.numeric(.), all_vars(!is.infinite(.))) %>% 
  summarise(yearly_growth_rate_dbh = dbh_diff/max_years,
            yearly_growth_rate_basal = basal_diff/max_years)  

all_gr %>% 
  group_by(forest_type) %>% 
  summarise(median(yearly_growth_rate_basal),
            median(yearly_growth_rate_dbh)) %>% 
  knitr::kable(digits = 2)
  
```

```{r}
all_gr %>% 
  group_by(forest_type) %>% 
  ggplot(aes(x = forest_type, y = yearly_growth_rate_basal)) +
  geom_jitter(alpha = 0.2, shape = 16) +
  geom_boxplot(aes(colour = forest_type), 
               outliers = FALSE, 
               linewidth = 0.8,
               fill = "transparent") +
  coord_cartesian(ylim = c(0, 10)) +
  theme(legend.position = "none") +
  ggtitle("Basal diameter") +
  
  all_gr %>% 
  group_by(forest_type) %>% 
  ggplot(aes(x = forest_type, y = yearly_growth_rate_dbh)) +
  geom_jitter(alpha = 0.2, shape = 16) +
  geom_boxplot(aes(colour = forest_type),
               outliers = FALSE, 
               linewidth = 0.8,
               fill = "transparent") +
  coord_cartesian(ylim = c(0, 10)) +
  theme(legend.position = "none") +
  ggtitle("DBH")
```

Yearly growth rate is generally slower in the secondary forest.

```{r}
fit_basal <- 
  lme4::lmer(log(dbase_mean) ~ days_num + cohort + (1|plant_id), 
             data = data)

fit_dbh <- 
  lme4::lmer(log(dbh_mean) ~ days_num + cohort + (1|plant_id), 
             data = data)

data %>% 
  drop_na(dbase_mean) %>% 
  ggplot(aes(x = years, y = log(dbase_mean), colour = cohort)) +
  geom_point(alpha = 0.3, size = 0.5, shape = 16) +
  geom_smooth(aes(y = predict(fit_basal), colour = cohort),
              method = "lm") +
  ggtitle("log basal diameter") +
  
  data %>% 
  drop_na(dbh_mean) %>% 
  ggplot(aes(x = years, y = log(dbh_mean), colour = cohort)) +
  geom_point(alpha = 0.3, size = 0.5, shape = 16) +
  geom_smooth(aes(y = predict(fit_dbh)),
              method = "lm") +
  ggtitle("log DBH") +
  
  patchwork::plot_layout(guides = "collect") +
  patchwork::plot_annotation(title = "geom_smooth(method = `lm`)") &
  theme(legend.position = "bottom")
```

```{r}
data %>% 
  drop_na(dbase_mean) %>% 
  ggplot(aes(x = years, y = log(dbase_mean), colour = cohort)) +
  geom_point(alpha = 0.3, size = 0.5, shape = 16) +
  geom_smooth(method = "gam") +
  ggtitle("log basal diameter") +
  
  data %>% 
  drop_na(dbh_mean) %>% 
  ggplot(aes(x = years, y = log(dbh_mean), colour = cohort)) +
  geom_point(alpha = 0.3, size = 0.5, shape = 16) +
  geom_smooth(method = "gam") +
  ggtitle("log DBH") +
  
  patchwork::plot_layout(guides = "collect") +
  patchwork::plot_annotation(title = "geom_smooth(method = `gam`)") &
  theme(legend.position = "bottom")
```

1^st^ cohort and 2^nd^ cohort of the secondary forest seedlings look very similar.
Perhaps some evidence for primary forest seedlings having initial higher growth rate?

We are seeing that drop in basal diameter for primary forest seedlings
simply because we stopped taking basal measurements for the last 2 surveys -
there are few data points, and those that do exist are likely there because
the seedling was too short to record DBH.

## Growth by species

```{r}
all_gr %>% 
  group_by(genus_species) %>% 
  summarise(median(yearly_growth_rate_dbh),
            median(yearly_growth_rate_basal)) %>% 
  arrange(desc(`median(yearly_growth_rate_dbh)`)) %>% 
  formattable::format_table(
    format = "html",
    digits = 2,
    list(`median(yearly_growth_rate_dbh)` = 
           formattable::color_tile("transparent", "lightpink"),
         `median(yearly_growth_rate_basal)` = 
           formattable::color_tile("transparent", "lightpink"))
  )
```

```{r}
#| fig-width: 7
#| fig-height: 7

all_gr %>% 
  group_by(genus_species) %>% 
  mutate(`median(yearly_growth_rate_basal)` = 
           median(yearly_growth_rate_basal)) %>% 
  ggplot(aes(x = genus_species,
             y = yearly_growth_rate_basal,
             group = genus_species)) +
  geom_jitter(alpha = 0.2, shape = 16) +
  geom_boxplot(aes(colour = `median(yearly_growth_rate_basal)`), 
               outliers = FALSE, 
               linewidth = 0.8,
               fill = "transparent") +
  coord_cartesian(ylim = c(0, 10)) +
  scale_colour_viridis_c() +
  theme(axis.title.x = element_blank()) +
  guides(x =  guide_axis(angle = 90)) +
  ggtitle("Basal diameter") +
  
  all_gr %>% 
  group_by(genus_species) %>% 
  mutate(`median(yearly_growth_rate_dbh)` = 
           median(yearly_growth_rate_dbh)) %>% 
  ggplot(aes(x = genus_species, 
             y = yearly_growth_rate_dbh)) +
  geom_jitter(alpha = 0.2, shape = 16) +
  geom_boxplot(aes(colour = `median(yearly_growth_rate_dbh)`),
               outliers = FALSE, 
               linewidth = 0.8,
               fill = "transparent") +
  coord_cartesian(ylim = c(0, 10)) +
  scale_colour_viridis_c() +
  theme(axis.title.x = element_blank()) +
  guides(x = guide_axis(angle = 90)) +
  ggtitle("DBH") +
  
  plot_layout(guides = "collect", ncol = 2) &
  theme(legend.position = "bottom") 
  
```

```{r}
#| fig-width: 7
#| fig-height: 10

all_gr %>% 
  group_by(genus_species) %>% 
  summarise(median(yearly_growth_rate_basal)) %>%
  right_join(data) %>% 
  drop_na(dbase_mean) %>% 
  ggplot(aes(x = years, y = log(dbase_mean), 
             group = genus_species)) +
  geom_point(alpha = 0.3, size = 0.5, shape = 16) +
  geom_smooth(aes(colour = `median(yearly_growth_rate_basal)`),
              linewidth = 1, method = "gam") +
  facet_wrap(~genus_species, ncol = 5) +
  scale_colour_viridis_c() +
  theme(legend.position = "none") +
  ggtitle("log basal diameter") +
  
  all_gr %>% 
  group_by(genus_species) %>% 
  summarise(median(yearly_growth_rate_dbh)) %>%
  right_join(data) %>% 
  drop_na(dbh_mean) %>% 
  ggplot(aes(x = years, y = log(dbh_mean),
         group = genus_species)) +
  geom_point(alpha = 0.3, size = 0.5, shape = 16) +
  geom_smooth(aes(colour = `median(yearly_growth_rate_dbh)`),
              linewidth = 1, method = "gam") +
  facet_wrap(~genus_species, ncol = 5) +
  scale_colour_viridis_c() +
  theme(legend.position = "none") +
  ggtitle("log DBH") +
  
  plot_layout(ncol = 1) &
  theme(legend.position = "bottom") 
```

A fast basal diameter growth rate doesn't always correlate to a fast DBH growth rate.

## Growth by species & forest type

Table of median yearly DBH growth rate in different forest types:

```{r}
all_gr %>% 
  group_by(genus_species, forest_type) %>% 
  summarise(yearly_growth_rate_dbh = median(yearly_growth_rate_dbh)) %>% 
  pivot_wider(names_from = forest_type, values_from = yearly_growth_rate_dbh) %>% 
  arrange(desc(primary)) %>% 
  mutate(diff = primary - secondary) %>% 
  formattable::format_table(
    format = "html",
    digits = 2,
    list(formattable::area(col = primary:secondary) ~ 
           formattable::color_tile("transparent", "lightpink"))
  ) 
```

Most species seem to grow faster in the primary forest.

```{r}
#| fig-width: 9
#| fig-height: 5

all_gr %>% 
  group_by(genus_species, forest_type) %>% 
  summarise(median(yearly_growth_rate_dbh)) %>%
  pivot_wider(names_from = "forest_type", 
              values_from = `median(yearly_growth_rate_dbh)`,
              id_cols = genus_species) %>% 
  ggplot(aes(y = primary, x = secondary, colour = genus_species)) +
  geom_point() +
  ggtitle("DBH - yearly growth rate") +
  
  all_gr %>% 
  group_by(genus_species, forest_type) %>% 
  summarise(median(yearly_growth_rate_basal)) %>%
  pivot_wider(names_from = "forest_type", 
              values_from = `median(yearly_growth_rate_basal)`,
              id_cols = genus_species) %>% 
  ggplot(aes(y = primary, x = secondary, colour = genus_species)) +
  geom_point() +
  ggtitle("basal diameter - yearly growth rate") +
  
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

Growth rate in primary and secondary forest don't correlate!

```{r}
#| fig-width: 7
#| fig-height: 10

data %>% 
  drop_na(dbase_mean) %>% 
  ggplot(aes(x = years, y = log(dbase_mean), colour = cohort)) +
  geom_point(alpha = 0.3, size = 0.5, shape = 16) +
  geom_smooth(method = "gam") +
  facet_wrap(~genus_species, ncol = 5) +
  xlim(0, 18) +
  ggtitle("log basal diameter") +
  
  data %>% 
  drop_na(dbh_mean) %>% 
  ggplot(aes(x = years, y = log(dbh_mean), colour = cohort)) +
  geom_point(alpha = 0.3, size = 0.5, shape = 16) +
  geom_smooth(method = "gam") +
  facet_wrap(~genus_species, ncol = 5) +
  ggtitle("log DBH") +
  
  plot_layout(guides = "collect", ncol = 1) &
  theme(legend.position = "bottom") 
  
```

## Climber-cut vs non-climber cut intensively sampled plots

Three of the secondary forest plots received a climber cutting treatment
(plots `05`, `11` and `14`) 
(see [O'Brien _et al_ 2019](https://doi.org/10.1111/1365-2664.13335)). 
Table of median yearly DBH growth rate:

```{r}
all_gr_pl <- 
  data %>% 
  filter(forest_type == "secondary") %>% 
  mutate(climber_cut = ifelse(
    plot %in% c("05", "11", "14"),
    TRUE, FALSE
  )) %>% 
  group_by(plant_id, genus_species, climber_cut) %>%
  summarise(dbh_diff = max(dbh_mean, na.rm = TRUE) - 
              min(dbh_mean, na.rm = TRUE),
            basal_diff = max(dbase_mean, na.rm = TRUE) - 
              min(dbase_mean, na.rm = TRUE),
            max_years = max(years, na.rm = TRUE), 
            .groups = "keep") %>% 
  filter_if(~is.numeric(.), all_vars(!is.infinite(.))) %>% 
  summarise(yearly_growth_rate_dbh = dbh_diff/max_years,
            yearly_growth_rate_basal = basal_diff/max_years)  

all_gr_pl %>% 
  group_by(genus_species, climber_cut) %>% 
  summarise(yearly_growth_rate_dbh = median(yearly_growth_rate_dbh)) %>% 
  pivot_wider(names_from = climber_cut, 
              values_from = yearly_growth_rate_dbh) %>% 
  rename(climbers_cut = `TRUE`, climbers_not_cut = `FALSE`) %>% 
  arrange(desc(climbers_not_cut)) %>% 
  mutate(diff = climbers_not_cut - climbers_cut) %>% 
  formattable::format_table(
    format = "html",
    digits = 2,
    list(formattable::area(col = climbers_not_cut:climbers_cut) ~ 
           formattable::color_tile("transparent", "lightpink"))
  )
```

The effect of climber cutting on growth seems to vary by species.

```{r}
#| fig-width: 7
#| fig-height: 10

all_gr_pl %>% 
  ggplot(aes(x = genus_species, 
             y = yearly_growth_rate_basal,
             colour = climber_cut)) +
  geom_jitter(alpha = 0.2, shape = 16) +
  geom_boxplot(outliers = FALSE, 
               linewidth = 0.8,
               fill = "transparent") +
  coord_cartesian(ylim = c(0, 10)) +
  theme(axis.title.x = element_blank()) +
  guides(x = guide_axis(angle = 90)) +
  ggtitle("Basal diameter") +
  
  all_gr_pl %>% 
  ggplot(aes(x = genus_species, 
             y = yearly_growth_rate_dbh,
             colour = climber_cut)) +
  geom_jitter(alpha = 0.2, shape = 16) +
  geom_boxplot(outliers = FALSE, 
               linewidth = 0.8,
               fill = "transparent") +
  coord_cartesian(ylim = c(0, 10)) +
  theme(axis.title.x = element_blank()) +
  guides(x = guide_axis(angle = 90)) +
  ggtitle("DBH") +
  
  plot_layout(guides = "collect",
              ncol = 1) &
  theme(legend.position = "bottom") 
  
```

Effect of climber cutting seems not so apparent when looking at the full distribution of the data. 
Perhaps more variation in growth rate in climber cut plots?

```{r}
#| fig-width: 7
#| fig-height: 10

data %>% 
  filter(forest_type == "secondary") %>% 
  mutate(climber_cut = ifelse(
    plot %in% c("05", "11", "14"),
    TRUE, FALSE
  )) %>% 
  drop_na(dbase_mean) %>% 
  ggplot(aes(x = years, y = log(dbase_mean))) +
  geom_point(alpha = 0.3, size = 0.5, shape = 16) +
  geom_smooth(aes(colour = climber_cut), size = 1, method = "gam") +
  facet_wrap(~plot, ncol = 3) +
  ggtitle("log basal diameter") +
  
  data %>% 
  filter(forest_type == "secondary") %>% 
  mutate(climber_cut = ifelse(
    plot %in% c("05", "11", "14"),
    TRUE, FALSE
  )) %>% 
  drop_na(dbase_mean) %>% 
  ggplot(aes(x = years, y = log(dbh_mean))) +
  geom_point(alpha = 0.3, size = 0.5, shape = 16) +
  geom_smooth(aes(colour = climber_cut), size = 1, method = "gam") +
  facet_wrap(~plot, ncol = 3) +
  ggtitle("log dbh")  +
  
  plot_layout(guides = "collect", ncol = 1) +
  plot_annotation(title = "Secondary forest seedlings by plot") &
  theme(legend.position = "bottom") 
  
```


