---
title: "Investigating size at planting/first survey"
author: 'eleanorjackson'
date: '`r format(Sys.time(), "%d %B, %Y")`' 
format: gfm+emoji
editor: source
execute:
  warning: false
---

```{r setup}
#| include: false

file_name <- knitr::current_input()

knitr::opts_chunk$set(
  fig.path =
    paste0("figures/", sub("\\.rmarkdown$", "", basename(file_name)), "/", sep = "")
)

ggplot2::theme_set(ggplot2::theme_classic(base_size = 10))
```

> Seedlings were not planted in the forest types at the same time,
> the secondary forest of SBE was planted first 
> then the left over seedlings in the nursery were planted in the 
> primary old growth forest of Danum Valley Conservation Area. Meaning… 
> seedlings grow for a different amount of time in the two forest types 
> and initial size is not equalized by randomization 
> (appears to differ on a species-by-species basis if I recall correctly)

Smaller seedlings usually have a faster rate of growth, 
which slows as they get larger.

Taking a look at this now that I've combined the different data sources
and done a little simple cleaning.
(see [00_combine-raw-data.R](code/scripts/00_combine-raw-data.R) and [01_clean-data.R](code/scripts/01_clean-data.R)).

```{r packages}
#| output: false

library("tidyverse")
library("here")
library("patchwork")
```

```{r}
data <- 
  readRDS(here::here("data", "derived", "data_cleaned.rds"))
```

Let's have a quick look at the timeline of surveys.

```{r}
#| fig-height: 3
#| fig-width: 7

data %>% 
  group_by(forest_type, survey_date) %>% 
  slice_head() %>% 
  ggplot(aes(y = forest_type, x = survey_date)) +
  geom_point(alpha = 0.6, shape = 16, size = 3)
```

We are waiting on the latest data (2024) from the SBE. 
__Do we have planting dates for the Danum Valley seedlings?__

Find the first survey date for each individual.

```{r}
data <- 
  data %>% 
  group_by(plant_id) %>%
  slice_min(survey_date, with_ties = FALSE) %>%
  ungroup() %>% 
  select(plant_id, survey_date) %>%
  rename(first_survey = survey_date) %>% 
  right_join(data) 
```

```{r}
data %>% 
  filter(survey_date == first_survey) %>% 
  summarise_at(c("dbase_mean", "dbh_mean"), 
               ~sum(is.na(.x))
               )
```

There are a lot of entries with a record for survival but no diameter measurement.
__Why would survival be recorded without a diameter measurement?__

```{r}
data %>% 
  filter(survey_date == first_survey) %>% 
  filter(is.na(dbase_mean) & is.na(dbh_mean)) %>% 
  group_by(forest_type) %>% 
  summarise(n())
```

This is mostly happening in the SBE data. 

```{r}
data %>% 
  filter(survey_date == first_survey) %>% 
  ggplot(aes(x = dbase_mean)) +
  geom_histogram(binwidth = 2) +
  facet_wrap(~forest_type, scales = "free_y") +
  xlim(0, 50) +
  
  data %>% 
  filter(survey_date == first_survey) %>% 
  ggplot(aes(x = dbh_mean)) +
  geom_histogram(binwidth = 2) +
  facet_wrap(~forest_type, scales = "free_y") +
  xlim(0, 50) +
  
  data %>% 
  filter(survey_date == first_survey) %>% 
  ggplot(aes(x = height_apex)) +
  geom_histogram(binwidth = 5) +
  facet_wrap(~forest_type, scales = "free_y") +
  xlim(0, 500) +
  
  plot_layout(ncol = 1)
```

Looks like it's going to be best to look at basal diameter. 
__Is there a way to estimate DBH from basal diameter or vice versa?__

```{r}
data %>% 
  filter(survey_date == first_survey) %>% 
  ggplot(aes(x = dbase_mean, 
             fill = forest_type,
             colour = forest_type)) +
  geom_density(alpha = 0.3) +
  facet_wrap(~genus_species) +
  stat_summary(aes(xintercept = ..x.., y = 0), 
               fun = mean, 
               geom = "vline", 
               orientation = "y",
               linetype = 2) +
  xlim(0, 20) +
  ggtitle("Size at first survey")
```

The primary forest (Danum) seedlings often have a greater spread of sizes and
a larger mean size in their first survey. 
They were planted later, but were larger at the time of planting.

But some species have pretty good overlap:
 
- _S. gibbosa_
- _S. macroptera_
- _S. faguetiana_

How about if we match the second survey of the primary forest to the first 
survey of the secondary forest?

```{r}
data <- 
  data %>% 
  filter(survey_date != first_survey) %>% 
  group_by(plant_id) %>%
  slice_min(survey_date, with_ties = FALSE) %>%
  ungroup() %>% 
  select(plant_id, survey_date) %>%
  rename(second_survey = survey_date) %>% 
  right_join(data) 
```

```{r}
data %>% 
  filter(case_when(
    forest_type == "primary" ~ survey_date == first_survey,
    forest_type == "secondary" ~ survey_date == second_survey,
  )) %>% 
  ggplot(aes(x = dbase_mean, 
             fill = forest_type,
             colour = forest_type)) +
  geom_density(alpha = 0.3) +
  facet_wrap(~genus_species) +
  stat_summary(aes(xintercept = ..x.., y = 0), 
               fun = mean, 
               geom = "vline", 
               orientation = "y",
               linetype = 2) +
  xlim(0, 20) 
```

That has closed the gap between means a little - 
for some species more than others.
It looks like there's better overlap in 
the range of sizes between the forest types.

Maybe there is more range in primary forest seedling size 
because of the two cohorts!

```{r}
data %>% 
  filter(
    forest_type == "primary" & 
      survey_date == first_survey &
      !is.na(old_new)) %>% 
  ggplot(aes(x = dbase_mean, 
             fill = forest_type,
             colour = forest_type,
             linetype = old_new)) +
  geom_density(alpha = 0.3) +
  facet_wrap(~genus_species) +
  stat_summary(aes(xintercept = ..x.., y = 0), 
               fun = mean, 
               geom = "vline", 
               orientation = "y") +
  xlim(0, 20) 
```

```{r}
data %>% 
  rowwise() %>% 
  mutate(type = paste(forest_type, old_new, sep = "_")) %>% 
  ungroup() %>% 
  filter(survey_date == first_survey) %>% 
  filter(case_when(
    forest_type == "primary" ~ !is.na(old_new),
    T ~ is.na(old_new)
  )) %>% 
  ggplot(aes(x = dbase_mean, 
             fill = type,
             colour = type)) +
  geom_density(alpha = 0.3) +
  facet_wrap(~genus_species) +
  stat_summary(aes(xintercept = ..x.., y = 0), 
               fun = mean, 
               geom = "vline", 
               orientation = "y",
               linetype = 2) +
  xlim(0, 20) 
```

The old and new secondary cohorts seem very close to each other
compared to the primary forest seedlings.

Biggest differences in old v new seen for 
_H. sangal, S. beccariana, S. faguetiana, S. macrophylla._

Now let's see if growth is different over time for these three groups.

```{r}
data %>%
  rowwise() %>% 
  mutate(type = paste(forest_type, old_new, sep = "_")) %>% 
  ungroup() %>% 
  ggplot(aes(y = dbase_mean,
             x = survey_date,
             fill = type,
             colour = type)) +
  geom_smooth() +
  facet_wrap(~genus_species) 
```

It looks like some species do better in the secondary forest 
and some in the primary, but perhaps the secondary forest seedlings are just ahead 
and growth rate is starting to level out?

Making that plot again but using `days_since_first_survey` on 
the y-axis, rather than `survey_date`.

```{r}
data <-
  data %>% 
  rowwise() %>% 
  mutate(
    days_since_first_survey =
      survey_date - first_survey
  ) %>% 
  ungroup()

data %>%
  rowwise() %>% 
  mutate(type = paste(forest_type, old_new, sep = "_")) %>% 
  ungroup() %>% 
  ggplot(aes(y = dbase_mean,
             x = days_since_first_survey,
             fill = type,
             colour = type)) +
  geom_smooth(alpha = 0.3) +
  facet_wrap(~genus_species, scales = "free_y") 
```

Looks like something weird is going on.

```{r}
data %>%
  rowwise() %>% 
  mutate(type = paste(forest_type, old_new, sep = "_")) %>% 
  ungroup() %>% 
  filter(type != "primary_NA") %>% 
  ggplot(aes(y = dbase_mean,
             x = days_since_first_survey,
             fill = type,
             colour = type)) +
  geom_point(size = 0.5, alpha = 0.3) +
  geom_smooth(alpha = 0.3) +
  facet_wrap(~genus_species, scales = "free_y") 
```

Seems like a few points are dragging down the secondary forest growth rate -
perhaps just smaller sample size as more trees have died.
