---
title: "Testing for spatial autocorrelation in survival"
author: 'eleanorjackson'
date: '`r format(Sys.time(), "%d %B, %Y")`' 
format: gfm+emoji
editor: source
execute:
  warning: false
---

```{r setup}
#| include: false

file_name <- knitr::current_input()

knitr::opts_chunk$set(
  fig.path =
    paste0("figures/", sub("\\.rmarkdown$", "", basename(file_name)), "/", sep = "")
)

set.seed(123)
ggplot2::theme_set(ggplot2::theme_classic(base_size = 10))
```

```{r}
#| output: FALSE

library("tidyverse")
library("here")
library("patchwork")
library("brms")
library("DHARMa")
library("DHARMa.helpers")
```

```{r}
data <- 
  readRDS(here::here("data", "derived", "data_cleaned.rds"))
```

# aggregate data for survival models

```{r}
data <-
  data %>%
  group_by(plant_id) %>%
  slice_min(survey_date, with_ties = FALSE) %>%
  select(plant_id, survey_date) %>%
  rename(first_survey = survey_date) %>%
  ungroup() %>% 
  right_join(data)

data <-
  data %>%
  rowwise() %>% 
  mutate(
    days =
      survey_date - first_survey) %>% 
  ungroup() %>% 
  mutate(years = as.numeric(days, units= "weeks")/52.25,
         days_num = as.numeric(days))
```

```{r}
interval_censored <-
  data %>% 
  filter(survival == 0) %>% 
  group_by(plant_id) %>% 
  slice_min(survey_date, with_ties = FALSE) %>% 
  ungroup() %>% 
  rename(time_to_dead = years) %>% 
  select(plant_id, genus_species, plot, forest_type, cohort, time_to_dead) %>% 
  mutate(censor = "interval")


interval_censored <-
  data %>% 
  filter(plant_id %in% interval_censored$plant_id) %>% 
  filter(survival == 1) %>% 
  group_by(plant_id) %>% 
  slice_max(survey_date, with_ties = FALSE) %>% 
  ungroup() %>% 
  rename(time_to_last_alive = years) %>% 
  select(plant_id, time_to_last_alive) %>% 
  right_join(interval_censored) 
  
  
right_censored <- 
  data %>% 
  filter(!plant_id %in% interval_censored$plant_id) %>% 
  group_by(plant_id) %>% 
  slice_max(survey_date, with_ties = FALSE) %>% 
  ungroup() %>% 
  rename(time_to_last_alive = years) %>% 
  select(plant_id, genus_species, plot, forest_type, 
         cohort, time_to_last_alive) %>% 
  mutate(censor = "right")

data_aggregated <- 
  bind_rows(interval_censored, right_censored) %>% 
  filter(time_to_last_alive > 0) %>% 
  mutate(plot_id = as.factor(paste(forest_type, plot, sep = "_")))

data_aggregated <- 
  data_aggregated %>% 
  mutate(censor_2 = case_when(
         censor == "right" ~ "right",
         censor == "interval" ~ "none")) %>% 
  mutate(time_to_dead = case_when(
    censor == "right" ~ time_to_last_alive,
    .default = time_to_dead
  )) %>% 
  rowwise() %>% 
  mutate(time_to_dead_2 = case_when(
    censor == "interval" ~ median(c(time_to_last_alive, time_to_dead)),
    .default = time_to_dead
  ) ) %>% 
  ungroup()
```

# generate grid

```{r}
x_vals <- 
  tibble(x = seq(from = 1, by = 10, length.out = 20),
       line = levels(filter(data, forest_type == "secondary")$line))

y_vals <- 
  tibble(y = seq(from = 1, by = 3, length.out = 110),
       position = 1:110) %>% 
  mutate(position = formatC(position,
                             width = 3,
                             format = "d",
                             flag = "0")) %>% 
  mutate(position = as.factor(position))

SBE_space <- 
  data %>% 
  filter(forest_type == "secondary") %>% 
  mutate(position = as.factor(position)) %>% 
  left_join(y_vals) %>% 
  left_join(x_vals) %>% 
  mutate(y = case_when(
    plot == "03" ~ y + 1000,
    plot == "05" ~ y + 500,
    plot == "08" ~ y + 700,
    plot == "11" ~ y + 1000,
    plot == "14" ~ y + 400,
    plot == "17" ~ y + 900,
  )) %>% 
  mutate(x = case_when(
    plot == "03" ~ x + 0,
    plot == "05" ~ x + 0,
    plot == "08" ~ x + 200,
    plot == "11" ~ x + 400,
    plot == "14" ~ x + 400,
    plot == "17" ~ x + 600,
  ))
```

```{r}
SBE_space %>% 
  filter(cohort == "1") %>% 
  select(plant_id, plot, x, y) %>% 
  distinct() %>% 
  ggplot(aes(x = x, y = y, colour = plot)) +
  geom_point(size = 0.5) 
```

```{r}
SBE_space_surv <- 
  SBE_space %>% 
  filter(cohort == "1") %>% 
  select(plant_id, plot, line, position, x, y) %>% 
  distinct() %>% 
  left_join(data_aggregated) %>% 
  drop_na() # NAs are left censored plants
```

# fit for just primary forest and cohort 1

```{r}
#| output: false

fit_secondary_part_pool <- 
  brm(time_to_last_alive|cens(x = censor, y2 = time_to_dead) ~ 
        1 + (1 | genus_species) + (1 | plot), 
      data = SBE_space_surv,
      family = brmsfamily("weibull"), 
      iter = 500,
      cores = 4,
      chains = 4,
      seed = 123,
      file = here::here("code", "notebooks", "models",
                        "2024-11-05_test-spatial-autocorrelation", 
                        "fit_weibull_sbe_part_pool.rds"),
      file_refit = "on_change")
```

```{r}
plot(fit_secondary_part_pool)
```

```{r}
mod_check <- dh_check_brms(fit_secondary_part_pool)
```

```{r}
#| fig-width: 10
#| fig-height: 10

testSpatialAutocorrelation(
  mod_check, 
  x = SBE_space_surv$x, 
  y = SBE_space_surv$y
)
```
_Plot showing each residual with at its x,y position, coloured according to the residual value. Residuals with 0.5 are coloured white, everything below 0.5 is coloured increasingly red, everything above 0.5 is coloured increasingly blue._

If you’ve modelled all the spatial auto-correlation in your data, 
then the Moran coefficient for your residuals will be slightly negative.

Fit without pooling of species rather than partial pooling.

```{r}
#| output: false

fit_secondary_no_pool <- 
  brm(time_to_last_alive|cens(x = censor, y2 = time_to_dead) ~ 
        1 + genus_species + (1 | plot), 
      data = SBE_space_surv,
      family = brmsfamily("weibull"), 
      iter = 500,
      cores = 4,
      chains = 4,
      seed = 123,
      file = here::here("code", "notebooks", "models",
                        "2024-11-05_test-spatial-autocorrelation", 
                        "fit_weibull_sbe_no_pool.rds"),
      file_refit = "on_change")
```

```{r}
plot(fit_secondary_no_pool)
```

```{r}
mod_check_no_pool <- dh_check_brms(fit_secondary_no_pool)
```

```{r}
#| fig-width: 10
#| fig-height: 10

testSpatialAutocorrelation(
  mod_check_no_pool, 
  x = SBE_space_surv$x, 
  y = SBE_space_surv$y
)
```
