---
title: "Diameter conversion"
author: 'eleanorjackson'
date: '`r format(Sys.time(), "%d %B, %Y")`' 
format: gfm+emoji
editor: source
execute:
  warning: false
---

```{r setup}
#| include: false

file_name <- knitr::current_input()

knitr::opts_chunk$set(
  fig.path =
    paste0("figures/", sub("\\.rmarkdown$", "", basename(file_name)), "/", sep = "")
)

ggplot2::theme_set(ggplot2::theme_classic(base_size = 10))
```

```{r packages}
#| output: false

library("tidyverse")
library("here")
library("patchwork")
```

Some of the tree diameter measurements were taken at breast height (1.3m),
but others were taken at at base of the tree (5 cm from the ground).

```{r}
data <- 
  readRDS(here::here("data", "derived", "data_cleaned.rds"))
```


```{r}
data %>% 
  ggplot(aes(x = dbase_mean)) +
  geom_histogram(binwidth = 2) +
  facet_wrap(~forest_type, scales = "free_y") +
  xlim(0, 50) +
  
  data %>% 
  ggplot(aes(x = dbh_mean)) +
  geom_histogram(binwidth = 2) +
  facet_wrap(~forest_type, scales = "free_y") +
  xlim(0, 50) +
  
  data %>% 
  ggplot(aes(x = height_apex)) +
  geom_histogram(binwidth = 5) +
  facet_wrap(~forest_type, scales = "free_y") +
  xlim(0, 500) +
  
  plot_layout(ncol = 1)
```

```{r}
data %>% 
  filter(is.na(dbh_mean)) %>% 
  ggplot(aes(x = dbase_mean)) +
  geom_histogram(binwidth = 2) +
  facet_wrap(~forest_type, scales = "free_y") +
  xlim(0, 50) +
  
  data %>% 
  filter(is.na(dbase_mean)) %>% 
  ggplot(aes(x = dbh_mean)) +
  geom_histogram(binwidth = 2) +
  facet_wrap(~forest_type, scales = "free_y") +
  xlim(0, 50)  +
  
  plot_layout(ncol = 1) +
  plot_annotation(title = "Cases where either basal or BH diameter is NA")
```

From [Philipson _et al._ 2020](https://doi.org/10.1126/science.aay4490),
we can calculate diameter at breast height from our `dbase_mean`
measurements with this equation:

$$
diameter_{1.3m} = \frac{DBH_{POM}}{exp(-0.029(POM - 1.3))}
$$


They used this when the point of measurement (POM) had to be raised above 
breast height (1.3m) due to buttress roots. 
I don't know if this applicable to our problem - 
our seedlings were too small for DBH to be taken so the POM is below 1.3m.

Also.. I think this would need to be species specific, 
and we'd be introducing a lot of error.

Perhaps if there is an approximate cut-off date for when trees are large
enough to be measured by DBH,
we can split the data into early and late growth and analyse separately.
Seedling to sapling, and sapling to adult? - different growth stages.

```{r}
data %>% 
  pivot_longer(c(dbase_mean, dbh_mean), 
               names_to = "measurement", 
               values_to = "size") %>% 
  drop_na(size) %>% 
  ggplot(aes(x = as.ordered(census_no), 
             fill = measurement)) +
  geom_bar() +
  facet_wrap(~forest_type, scales = "free_y", ncol = 1)
```

Let's make the same figure but as proportions.

```{r}
data %>% 
  pivot_longer(c(dbase_mean, dbh_mean), 
               names_to = "measurement", 
               values_to = "size") %>% 
  drop_na(size) %>% 
  ggplot(aes(x = as.ordered(census_no), 
             fill = measurement)) +
  geom_bar(position = "fill") +
  facet_wrap(~forest_type, scales = "free_y", ncol = 1)
```

The SBE census 8 shows quite a big (proportional) drop in DBH measurements?

From eyeballing it, from census 6 in the primary forest and census 10 
in the secondary forest, we seem to have a ~50:50 split in measurements.

Species grow at different rates, 
so perhaps we'll see differences there.

```{r}
#| fig-width: 8
#| fig-height: 7

data %>% 
  filter(forest_type == "primary") %>% 
  pivot_longer(c(dbase_mean, dbh_mean), 
               names_to = "measurement", 
               values_to = "size") %>% 
  drop_na(size) %>% 
  ggplot(aes(x = as.ordered(census_no), 
             fill = measurement)) +
  geom_bar(position = "fill") +
  facet_wrap(~genus_species, ncol = 3,
             axis.labels = "all_x", axes = "all_x") +
  ggtitle("Primary forest") +
  guides(x =  guide_axis(angle = 90)) +
  theme(legend.position = "top", legend.justification = "left")
```

Basal diameter measurements are used for longer in _D. conformis_ and _S. gibbosa_.
They have 100% basal measurements till census 5/6, 
whereas other sp. tend to only have 100% basal measurements till survey 3.

Draw again for the secondary forest seedlings.

```{r}
#| fig-width: 8
#| fig-height: 7

data %>% 
  filter(forest_type == "secondary") %>% 
  pivot_longer(c(dbase_mean, dbh_mean), 
               names_to = "measurement", 
               values_to = "size") %>% 
  drop_na(size) %>% 
  ggplot(aes(x = as.ordered(census_no), 
             fill = measurement)) +
  geom_bar(position = "fill") +
  facet_wrap(~genus_species, ncol = 3,
             axis.labels = "all_x", axes = "all_x") +
  ggtitle("Secondary forest") +
  guides(x =  guide_axis(angle = 90)) +
  theme(legend.position = "top", legend.justification = "left")

```

I'm not really seeing that same pattern in the SBE data - 
fairly similar across species?


```{r}
data %>% 
  pivot_longer(c(dbase_mean, dbh_mean), 
               names_to = "measurement", 
               values_to = "size") %>% 
  drop_na(size) %>% 
  group_by(forest_type, plant_id, census_no) %>% 
  summarise(n_measurements_per_plant = n()) %>% 
  group_by(n_measurements_per_plant, forest_type) %>% 
  summarise(n()) 
```

```{r}
data %>% 
  pivot_longer(c(dbase_mean, dbh_mean), 
               names_to = "measurement", 
               values_to = "size") %>% 
  drop_na(size) %>% 
  group_by(forest_type, plant_id, census_no) %>% 
  summarise(n_measurements_per_plant = n()) %>% 
  group_by(n_measurements_per_plant, forest_type, census_no) %>% 
  summarise(n_plants = n()) %>% 
  ggplot(aes(x = census_no, y = n_plants, 
             fill = as.ordered(n_measurements_per_plant))) +
  geom_col(position = "fill") +
  facet_wrap(~forest_type, ncol = 1,
             axis.labels = "all_x", axes = "all_x") +
  guides(x =  guide_axis(angle = 90)) +
  theme(legend.position = "top", legend.justification = "left") +
  geom_hline(yintercept = 0.5, colour = "white", linetype = 2) +
  geom_hline(yintercept = 0.75, colour = "white", linetype = 3)

```

This figure shows us how many plants have *both* a DBH and basal diameter
measurement in each survey.

Looking at this, I want to say from census 6 onwards we should use DBH,
but there is that dip in the SBE data with censuses 7, 8, 9, 
having many trees with only a basal diameter measurement.

We also don't yet how we're going to line up the 2 forest types in terms of 
dates / size, since the SBE seedlings were planted & surveyed earlier and 
were larger at the time of their first survey (see [2024-08-07_investigate-initial-size.md](2024-08-07_investigate-initial-size.md)).

```{r}
data %>% 
  filter(forest_type == "secondary" &
           census_no == "08") %>% 
  group_by(census_id, plot) %>% 
  summarise(n_distinct(plant_id))
```

Census 8 of the SBE is a census of the intensive plots - 
so covers all 6 plots. Don't know why basal diameters were used.

Update: I've had an idea about why basal diameters were used in census 8,
there were two cohorts!! 
The new cohort of seedlings would have been too small for DBH.

```{r}
#| fig-height: 6
#| fig-width: 7

data %>% 
  mutate(cohort = ifelse(forest_type == "secondary",
                         paste(forest_type, old_new, sep = "_"),
                         forest_type)) %>% 
  filter(cohort != "secondary_NA" ) %>% 
  pivot_longer(c(dbase_mean, dbh_mean), 
               names_to = "measurement", 
               values_to = "size") %>% 
  drop_na(size) %>% 
  group_by(cohort, plant_id, census_no) %>% 
  summarise(n_measurements_per_plant = n()) %>% 
  group_by(n_measurements_per_plant, cohort, census_no) %>% 
  summarise(n_plants = n()) %>% 
  ggplot(aes(x = census_no, y = n_plants, 
             fill = as.ordered(n_measurements_per_plant))) +
  geom_col(position = "fill") +
  facet_wrap(~cohort, ncol = 1,
             axis.labels = "all_x", axes = "all_x") +
  guides(x =  guide_axis(angle = 90)) +
  theme(legend.position = "top", legend.justification = "left") +
  geom_hline(yintercept = 0.5, colour = "white", linetype = 2) +
  geom_hline(yintercept = 0.75, colour = "white", linetype = 3)

```

Hmm looks better but census 8 is still odd in the old SBE cohort.
