---
title: "Cleaning census numbers"
author: 'eleanorjackson'
date: '`r format(Sys.time(), "%d %B, %Y")`' 
format: gfm+emoji
editor: source
execute:
  warning: false
---

```{r setup}
#| include: false

file_name <- knitr::current_input()

knitr::opts_chunk$set(
  fig.path =
    paste0("figures/", sub("\\.rmarkdown$", "", basename(file_name)), "/", sep = "")
)

ggplot2::theme_set(ggplot2::theme_classic(base_size = 10))
```

```{r packages}
#| output: false

library("tidyverse")
library("here")
library("patchwork")
```

Following up on 
[2024-08-28_cleaning-census-no.md](code/notebooks/2024-08-28_cleaning-census-no.md).

I've realised that in the SBE data `sample` is nested within `data_origin`. :facepalm:

```{r}
data <- 
  readRDS(here::here("data", "derived", "data_cleaned.rds"))
```

I've created a `census_no` variable 
which orders the censuses by median survey date.

```{r}
data %>% 
  filter(forest_type == "secondary") %>% 
  group_by(census_id, census_no) %>% 
  summarise(median_date = median(survey_date, na.rm = TRUE),
            min_date = min(survey_date, na.rm = TRUE),
            max_date = max(survey_date, na.rm = TRUE),
            days = max_date - min_date
            ) %>% 
  arrange(max_date) %>% 
  knitr::kable()
```

Some of the censuses took a long time to complete - 
suggests we should use date rather than census or year when modelling.

```{r}
data %>% 
  filter(forest_type == "secondary") %>% 
  group_by(census_id) %>% 
  summarise(n_distinct(plot)) %>% 
  knitr::kable()
```

There are 6 plots. 
Not every census surveys all 6 plots.

```{r}
data_complete_censuses <-
  data %>% 
  filter(forest_type == "secondary") %>% 
  filter(str_detect(census_id, "full_measurement") |
           str_detect(census_id, "intensive")) %>% 
  filter(!census_id == "intensive_04") 
```


```{r}
data_complete_censuses %>% 
  filter(forest_type == "secondary") %>% 
  mutate(month = lubridate::month(survey_date)) %>% 
  group_by(survey_date) %>% 
  slice_head() %>% 
  ggplot(aes(y = forest_type, x = survey_date, colour = as.ordered(census_no))) +
  geom_point(alpha = 0.6, shape = "|", size = 7) +
  theme_bw() +
  scale_x_date(minor_breaks = "year") +
  scale_colour_viridis_d() +
  theme(legend.position = "none") +
  ggtitle("Complete censuses") +
  
  data %>% 
  filter(forest_type == "secondary") %>% 
  mutate(month = lubridate::month(survey_date)) %>% 
  group_by(survey_date) %>% 
  slice_head() %>% 
  ggplot(aes(y = forest_type, x = survey_date, colour = as.ordered(census_no))) +
  geom_point(alpha = 0.6, shape = "|", size = 7) +
  theme_bw() +
  scale_x_date(minor_breaks = "year") +
  scale_colour_viridis_d() +
  theme(legend.position = "none") +
  ggtitle("All censuses") +
  
  patchwork::plot_layout(ncol = 1)
  
```

```{r}
data %>% 
  filter(forest_type == "secondary") %>%
  group_by(plant_id, census_no) %>% 
  summarise(n()) %>% 
  filter(`n()` >1) %>% 
  knitr::kable()
```

Duplicate trees within a single census are mostly trees without a "position",
so they've ended up with the same `plant_id` as another individual.

