---
title: "Single forest growth models"
author: 'eleanorjackson'
date: '`r format(Sys.time(), "%d %B, %Y")`' 
format: 
  gfm:
    toc: true
editor: source
execute:
  warning: false
---

```{r setup}
#| include: false

file_name <- knitr::current_input()

knitr::opts_chunk$set(
  fig.path =
    paste0("figures/", sub("\\.rmarkdown$", "", basename(file_name)), 
           "/", sep = "")
)

set.seed(123)
ggplot2::theme_set(ggplot2::theme_bw(base_size = 10))
```

I fit some models with _either_ the Danum Valley (old-growth, DV) _or_
Sabah Biodiversity Experiment (logged, SBE) data, 
in order to look at how growth might respond to 
some variables which are only relevant to one forest type.

```{r}
library("tidyverse")
library("tidybayes")
library("brms")
```

```{r}
mod_clim <- 
  readRDS(here::here("output", 
          "models", 
          "grow-extra-noft", 
          "growth_model_climber.rds"))

mod_cohort <- 
  readRDS(here::here("output", 
          "models", 
          "grow-extra-noft", 
          "growth_model_cohort.rds"))

mod_canopy <- 
  readRDS(here::here("output", 
          "models", 
          "grow-extra-noft", 
          "growth_model_canopy.rds"))
```

## SBE - climber cut vs non-climber cut 

```{r}
plot(mod_clim,
     variable = "^b_",
     ask = FALSE,
     regex = TRUE,
     newpage = FALSE,
     nvariables = 2)
```

Half of the six intensively censused SBE plots recieved
a climber cutting treatment 
(see [O’Brien *et al* 2019](https://doi.org/10.1111/1365-2664.13335)).

Estimated parameters look very similar between groups.
Perhaps _k_ a little higher for seedlings in climber-cut plots.

## SBE - cohort one vs cohort two

```{r}
plot(mod_cohort,
     variable = "^b_",
     ask = FALSE,
     regex = TRUE,
     newpage = FALSE,
     nvariables = 2)
```

Seedlings in cohort one were allegedly older at the time of planting 
out into the forest, 
but the two groups actually had a similar initial size, 
see [2024-08-07_investigate-initial-size](2024-08-07_investigate-initial-size.md). Likely however that the first cohort of seedlings were more pot-bound.

_A_ is higher for the 1^st^ cohort and
_k_ looks similar between groups.
Seedlings in the 2^nd^ cohort reach max RGR a little faster 
than seedlings in the 1^st^ cohort (_delay_).


## DV - understory vs gap

```{r}
plot(mod_canopy,
     variable = "^b_",
     ask = FALSE,
     regex = TRUE,
     newpage = FALSE,
     nvariables = 2)
```

In DV we have paired canopy gap and understory plots.
(Although I only use the gap plots in this project.)

This one did not fit so well.
I got some Rhat warnings and you can see the chains didn't mix well
for _delay_.

_k_ and _A_ both look higher in the understory, 
which is counter-intuitive and not what the data looks like, 
see [2025-05-28_understory-plots](2025-05-28_understory-plots.md).
