---
title: "Descriptive plots for Maria"
author: 'eleanorjackson'
date: '`r format(Sys.time(), "%d %B, %Y")`' 
format: pdf
editor: source
execute:
  warning: false
---

```{r setup}
#| include: false

file_name <- knitr::current_input()

knitr::opts_chunk$set(
  fig.path =
    paste0("figures/", sub("\\.rmarkdown$", "", basename(file_name)), 
           "/", sep = "")
)

set.seed(123)
ggplot2::theme_set(ggplot2::theme_classic(base_size = 10))
```

```{r}
#| echo: false

library("tidyverse")
library("here")
library("patchwork")
```

```{r}
#| echo: false

data <- 
  readRDS(here::here("data", "derived", "data_cleaned.rds")) 
```

## Initial size

The secondary forest seedlings (SBE) were planted first and
the leftovers were planted in the primary forest (Danum) at a later date.

In the below figure we can see that the primary forest seedlings were
smaller than the secondary forest seedlings at the time of their first survey,
and have a smaller range of sizes.
This could be beacuse they were in pots for longer causing their growth to be stunted.

This effects some species more than others.

```{r}
#| echo: false

data <- 
  data %>% 
  drop_na(dbase_mean) %>% 
  group_by(plant_id) %>%
  slice_min(survey_date, with_ties = FALSE) %>%
  ungroup() %>% 
  select(plant_id, survey_date) %>%
  rename(first_survey = survey_date) %>% 
  right_join(data) 
```

```{r}
#| echo: false
#| fig.height: 6

data %>% 
  rowwise() %>% 
  mutate(type = paste(forest_type, cohort, sep = "_")) %>% 
  ungroup() %>% 
  filter(survey_date == first_survey) %>% 
  ggplot(aes(x = dbase_mean, 
             fill = type,
             colour = type)) +
  geom_density(alpha = 0.3) +
  facet_wrap(~genus_species) +
  stat_summary(aes(xintercept = ..x.., y = 0), 
               fun = mean, 
               geom = "vline", 
               orientation = "y",
               linetype = 2) +
  xlim(0, 25) +
  theme(legend.position = "bottom") +
  xlab("Basal diameter") +
  ggtitle("Size at first survey")
```

```{r}
#| echo: false

data %>% 
  filter(survey_date == first_survey) %>% 
  group_by(genus_species, forest_type, cohort) %>% 
  summarise(`median_basal_diameter` = median(dbase_mean, na.rm = TRUE),
           IQR = IQR(dbase_mean, na.rm = TRUE)) %>% 
  knitr::kable()
```


## Irregular & unequal sampling effort

In the below figure each line represents a date on which data was collected
and the lines are coloured by the census number. 
Each census took multiple days (or months in the case of the SBE) to complete.

```{r}
#| echo: false
#| fig_width: 9

data %>% 
  mutate(month = lubridate::month(survey_date)) %>% 
  group_by(survey_date) %>% 
  slice_head() %>% 
  ggplot(aes(y = forest_type, x = survey_date, 
             colour = census_no)) +
  geom_point(alpha = 0.6, shape = "|", size = 10) +
  theme_bw() +
  scale_x_date(minor_breaks = "year") +
  scale_colour_viridis_d() +
  theme(legend.position = "none")
```


```{r}
#| echo: false

data %>% 
  select(census_no, plot, forest_type) %>%
  distinct() %>% 
  ggplot(aes(x = census_no, y = plot)) +
  geom_point(shape = 16, alpha = 0.8) +
  facet_wrap(~forest_type, scales = "free_x") +
  guides(x =  guide_axis(angle = 90)) 
```

There have been 22 censuses of the primary forest seedlings, and 26 censuses of the secondary forest seedlings.

Plot 17 in the secondary forest was not sampled as frequently as the other secondary forest plots as it wasn’t included in surveys for the climber cutting study.

## Climber cutting

Three of the secondary forest plots received a climber cutting treatment (plots 05, 11 and 14) 
(see [O’Brien et al 2019](https://doi.org/10.1111/1365-2664.13335)). 
When data was collected for this study only two planting lines 
from each of the climber cut (plots 05, 11 and 14) and 
control plots (plots 03 and 08) were surveyed 
since a full census would take too long.
