---
title: "TRY traits"
author: 'eleanorjackson'
date: '`r format(Sys.time(), "%d %B, %Y")`' 
format: 
  gfm:
    toc: true
editor: source
execute:
  warning: false
---

```{r setup}
#| include: false

file_name <- knitr::current_input()

knitr::opts_chunk$set(
  fig.path =
    paste0("figures/", sub("\\.rmarkdown$", "", basename(file_name)), 
           "/", sep = "")
)

set.seed(123)
ggplot2::theme_set(ggplot2::theme_bw(base_size = 10))
```

Which of our species are in the TRY database?

```{r}
library("tidyverse")
library("tidybayes")
```

```{r}
try_sp <- 
  read_tsv(here::here("data", "raw", "TryAccSpecies.txt")) %>% 
  mutate(genus_species = str_replace(AccSpeciesName, " ", "_")) 
```

```{r}
our_sp <- readRDS(here::here("data", "derived", "data_cleaned.rds")) %>% 
  select(genus_species) %>% 
  distinct()

our_sp
```


```{r}
inner_join(our_sp, try_sp) %>% glimpse
```

One of our species is not in the TRY database: `Parashorea_tomentella`

```{r}
inner_join(our_sp, try_sp) %>% 
  pull(AccSpeciesID) 
```

the `AccSpeciesID`'s we want:

49981, 49988, 19393, 49960, 50006, 49952, 49933, 18749, 49987, 40272, 49927, 50001, 29483, 49973

and the trait IDs:

- 4083	Leaf area per leaf dry mass (specific leaf area, SLA or 1/LMA) of total leaf area	
- 3086	Leaf area per leaf dry mass (specific leaf area, SLA or 1/LMA) petiole, rhachis and midrib excluded
- 3115	Leaf area per leaf dry mass (specific leaf area, SLA or 1/LMA): petiole excluded	
- 3116	Leaf area per leaf dry mass (specific leaf area, SLA or 1/LMA): petiole included	
- 3117	Leaf area per leaf dry mass (specific leaf area, SLA or 1/LMA): undefined if petiole is in- or excluded
- 4	Stem specific density (SSD, stem dry mass per stem fresh volume) or wood density
