---
title: "Time to inflection of the growth curve"
author: 'eleanorjackson'
date: '`r format(Sys.time(), "%d %B, %Y")`' 
format: 
  gfm:
    toc: true
editor: source
execute:
  warning: false
---

```{r setup}
#| include: false

file_name <- knitr::current_input()

knitr::opts_chunk$set(
  fig.path =
    paste0("figures/", sub("\\.rmarkdown$", "", basename(file_name)), 
           "/", sep = "")
)

set.seed(as.numeric(Sys.Date()))
ggplot2::theme_set(ggplot2::theme_bw(base_size = 13))
```

Sometimes _Ti_ is before the first measurement of the seedling in the field.

```{r}
library("tidyverse")
library("tidybayes")
library("brms")
library("modelr")
library("patchwork")
library("ggtext")
```

```{r}
mod_gro <-
  readRDS(here::here("output", "models",
                     "growth_model_base_p3_allo.rds"))
```

```{r}

data <-
  readRDS(here::here("data", "derived", "data_cleaned.rds"))

alive_trees <-
  data %>%
  filter(survival == 1) %>%
  filter(! if_all(c(dbh_mean, dbase_mean), is.na))

well_sampled_alive_trees <-
  alive_trees %>%
  group_by(plant_id) %>%
  summarise(n = n()) %>%
  filter(n > 2)

data_gro <-
  alive_trees %>%
  filter(plant_id %in% well_sampled_alive_trees$plant_id)
```

```{r}
delay_ests <-
  gather_draws(
    model = mod_gro,
    b_delay_forest_typeprimary,
    b_delay_forest_typelogged) %>%
  mutate(forest_type = case_when(
    grepl("logged", .variable) ~ "SBE",
    grepl("primary", .variable) ~ "Danum")) 
```

```{r}
ggplot(delay_ests,
       aes(x = .value, fill =forest_type)) +
  geom_density(alpha = 0.4) +
  xlab("Years")
```

```{r}
delay_ests %>% 
  group_by(forest_type) %>% 
  summarise(median(.value))
```

Time zero is the first survey for an individual seedling.

- For SBE cohort one, planting was between July 2002 and September 2003 and first survey was between January 2004 and April 2004
- For SBE cohort two, planting was between January 2009 and October 2010 and first survey was in May 2011
- For Danum seedlings, planting took place at the same time as censusing, between November 2004 and February 2005

For the first cohort of seedlings planted into SBE
between 18 and 20 months elapsed between planting and censusing (i.e, 1.5 to 1.7 years).

For the second cohort between 7 and 28 months
elapsed between planting and censusing (i.e, 0.6 to 2.3 years).

So, our estimates of _Ti_ (time to inflection of the growth curve) for SBE seedlings
fit within this range (-0.6 to -2.3).

Danum seedlings were censused at planting, but _Ti_ is still negative - 
suggests they reached max growth rates before planting into the field.

