---
title: "Diameter conversion"
author: 'eleanorjackson'
date: '`r format(Sys.time(), "%d %B, %Y")`' 
format: gfm+emoji
editor: source
execute:
  warning: false
---

```{r setup}
#| include: false

file_name <- knitr::current_input()

knitr::opts_chunk$set(
  fig.path =
    paste0("figures/", sub("\\.rmarkdown$", "", basename(file_name)), "/", sep = "")
)

ggplot2::theme_set(ggplot2::theme_classic(base_size = 10))
```

```{r packages}
#| output: false

library("tidyverse")
library("here")
library("patchwork")
```

Some of the tree diameter measurements were taken at breast height (1.3m),
but others were taken at at base of the tree (5 cm from the ground).

```{r}
data <- 
  readRDS(here::here("data", "derived", "data_cleaned.rds"))
```


```{r}
data %>% 
  ggplot(aes(x = dbase_mean)) +
  geom_histogram(binwidth = 2) +
  facet_wrap(~forest_type, scales = "free_y") +
  xlim(0, 50) +
  
  data %>% 
  ggplot(aes(x = dbh_mean)) +
  geom_histogram(binwidth = 2) +
  facet_wrap(~forest_type, scales = "free_y") +
  xlim(0, 50) +
  
  data %>% 
  ggplot(aes(x = height_apex)) +
  geom_histogram(binwidth = 5) +
  facet_wrap(~forest_type, scales = "free_y") +
  xlim(0, 500) +
  
  plot_layout(ncol = 1)
```

```{r}
data %>% 
  filter(is.na(dbh_mean)) %>% 
  ggplot(aes(x = dbase_mean)) +
  geom_histogram(binwidth = 2) +
  facet_wrap(~forest_type, scales = "free_y") +
  xlim(0, 50) +
  
  data %>% 
  filter(is.na(dbase_mean)) %>% 
  ggplot(aes(x = dbh_mean)) +
  geom_histogram(binwidth = 2) +
  facet_wrap(~forest_type, scales = "free_y") +
  xlim(0, 50)  +
  
  plot_layout(ncol = 1) +
  plot_annotation(title = "Cases where either basal or BH diameter is NA")
```

From [Philipson _et al._ 2020](https://doi.org/10.1126/science.aay4490),
we can calculate diameter at breast height from our `dbase_mean`
measurements with this equation:

$$
diameter_{1.3m} = \frac{DBH_{POM}}{exp(-0.029(POM - 1.3))}
$$


They used this when the point of measurement (POM) had to be raised above 
breast height (1.3m) due to buttress roots. 
I don't know if this applicable to our problem - 
our seedlings were too small for DBH to be taken so the POM is below 1.3m.

Also.. I think this needs to be species specific.
